---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# causalDisco <a href="https://bjarkehautop.github.io/causalDisco/"><img src="man/figures/logo.svg" align="right" height="139" alt="causalDisco website" /></a>

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/BjarkeHautop/causalDisco/graph/badge.svg)](https://app.codecov.io/gh/BjarkeHautop/causalDisco)
[![R-CMD-check](https://github.com/BjarkeHautop/causalDisco/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/BjarkeHautop/causalDisco/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/causalDisco)](https://CRAN.R-project.org/package=causalDisco)
<!-- badges: end -->

causalDisco provides a unified interface for causal discovery on observational data.
It wraps multiple causal discovery backends under a common, consistent syntax. 

## Motivation

Causal discovery methods exist in many ecosystems, for example in bnlearn, pcalg, or Tetrad, but their APIs vary widely.

causalDisco unifies them under one clear grammar, making it easy to compare results, switch algorithms, and focus on
scientific questions rather than package quirks.

Time to hit the disco 🪩

## Installation 

### Install causalDisco

To install causalDisco ensure you first have installed Rust as described below.

Then you can install the development version of causalDisco from GitHub using pak:

```{r install, eval=FALSE}
pak::pkg_install("https://github.com/BjarkeHautop/causalDisco")
```

or with all suggested packages (note that this requires a valid Java / JDK installation for rJava as described below):

```{r install w depend, eval=FALSE}
pak::pkg_install("https://github.com/BjarkeHautop/causalDisco", dependencies = TRUE)
```

### Installing Rust

causalDisco depends on the package [caugi](https://github.com/frederikfabriciusbjerre/caugi), which requires Rust to
be installed on your system. See https://rust-lang.org/tools/install/ for instructions on how to
install Rust.

### Installing Java / JDK

causalDisco provides an interface to the Java library [Tetrad](https://github.com/cmu-phil/tetrad) for
causal discovery algorithms. To use algorithms from Tetrad you need to install a Java Development Kit (JDK) >= 21.
We recommend Eclipse Temurin (OpenJDK), available at https://adoptium.net for all major operating systems.

Alternatively, we provide a helper function to install Temurin JDK 25 on macOS and Windows:

```{r install java, eval=FALSE}
causalDisco::install_java()
```

The current supported version of Tetrad can then be installed by calling

```{r install tetrad, eval=FALSE}
causalDisco::install_tetrad()
```

To verify everything is set up correctly you can run `check_tetrad_install()`:

```{r check tetrad install}
causalDisco::check_tetrad_install()
```

## Example

With causalDisco you can currently run causal discovery algorithms from the package causalDisco itself,
the Java library Tetrad, the R package bnlearn, and the R package pcalg.

```{r example}
library(causalDisco)

# Load data
data(tpc_example)

# Define background tiered knowledge object
kn <- knowledge(
  tpc_example,
  tier(
    child ~ starts_with("child"),
    youth ~ starts_with("youth"),
    old ~ starts_with("old")
  )
)

# Use Tetrad PC algorithm with conditional Gaussian test
# Requires Tetrad to be installed
if (check_tetrad_install()$installed && check_tetrad_install()$java_ok) {
  tetrad_pc <- pc(
    engine = "tetrad", # Use the Tetrad implementation
    test = "conditional_gaussian", # Use conditional Gaussian test
    alpha = 0.05 # Significance level for the test
  )
  disco_tetrad_pc <- disco(data = tpc_example, method = tetrad_pc, knowledge = kn)
}

# Use causalDisco's own implementation of the tges algorithm with temporal BIC score
cd_tges <- tges(
  engine = "causalDisco", # Use the causalDisco implementation
  score = "tbic" # Use temporal BIC score
)
disco_cd_tges <- disco(data = tpc_example, method = cd_tges, knowledge = kn)
```

You can visualize the resulting causal graph using the `plot()` function:

```{r plot}
#| fig.alt: "A causal graph with the known tiers indicated by vertical positioning of the nodes."
plot(disco_cd_tges)
```

## Questions

- Updated all the vignettes - anything missing?

## TODO

- Register a custom edge type for caugi to represent forbidden edges differently than normal directed edges.
Would simplify plotting logic and easier to extend later.

```{r register edge, eval = FALSE}
# Doesn't work atm because CRAN version requires length 3 (will work in next release)
caugi::register_caugi_edge(
  glyph = "!-->",
  tail_mark = "arrow",
  head_mark = "tail",
  class = "directed",
  symmetric = FALSE
)
```

### CRAN TODO

- Update Description: Gave a suggested DESCRIPTION.

- Change example in README to not require Tetrad?

- Update hexsticker logo.

- Push to discocoders and update the links, pkgdown, ...

## Bugs & requests

Bug reports and feature requests are welcome:

[open an issue](https://github.com/BjarkeHautop/causalDisco/issues).
