---
title: "Integrating custom tests and scores into causal discovery algorithms"
vignette: >
  %\VignetteIndexEntry{Integrating custom tests and scores into causal discovery algorithms}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    fig.align: "center"
---

```{r}
#| label: setup
library(causalDisco)
```

This vignette illustrates how to use custom conditional independence tests and
custom scores with the `disco()` function in causalDisco. We show how to define
user-specified tests and scores, and how to run causal discovery using these
custom functions.

Note that Tetrad-based algorithms do not currently support custom CI tests, so
this vignette focuses on R-based algorithms. By the end, you will be able to
seamlessly integrate your own evaluation functions into causalDisco workflows.

# Custom Tests

The interface for custom tests depends on the engine being used. Below, we
provide examples for both the causalDisco and pcalg engines, as well as bnlearn.

## causalDisco and pcalg engines

causalDisco and pcalg engines allow users to specify custom tests by providing a
function with the signature `function(x, y, conditioning_set, suff_stat)`, where
`x` and `y` are the variables being tested for independence, `conditioning_set`
is the conditioning set, and `suff_stat` is a list of sufficient statistics
needed for the test (e.g. correlation matrix and sample size). The function
should return a p-value indicating whether `x` and `y` are conditionally
independent given `conditioning_set`.

Of course the `suff_stat` argument could just be the raw data, but for
efficiency reasons, it is often better to precompute some sufficient statistics.

An example of a custom CI test is a partial correlation test based on the Fisher
Z-transform, which can be implemented as follows:

```{r}
#| label: my_custom_test
my_test <- function(x, y, conditioning_set, suff_stat) {
  C <- suff_stat$C
  n <- suff_stat$n

  vars <- c(x, y, conditioning_set)
  C_sub <- C[vars, vars, drop = FALSE]
  K <- solve(C_sub)
  r <- -K[1, 2] / sqrt(K[1, 1] * K[2, 2])
  z <- 0.5 * log((1 + r) / (1 - r))

  stat <- sqrt(n - length(conditioning_set) - 3) * abs(z)

  pval <- 2 * (1 - pnorm(stat))

  pval
}

my_suff_stat <- function(data) {
  list(
    C = cor(data),
    n = nrow(data)
  )
}
```

This can then be used with the causalDisco engine as follows:

```{r}
#| label: using_my_custom_test_causalDisco
data(tpc_example)

my_tpc <- tpc(
  engine = "causalDisco",
  test = my_test,
  alpha = 0.05,
  suff_stat_fun = my_suff_stat
)
result <- disco(data = tpc_example, method = my_tpc)
plot(result)
```

or the pcalg engine as follows:

```{r}
#| label: using_my_custom_test_pcalg
my_pc <- pc(
  engine = "pcalg",
  test = my_test,
  alpha = 0.05,
  suff_stat_fun = my_suff_stat
)
result <- disco(data = tpc_example, method = my_pc)
plot(result)
```

If the custom test requires additional arguments, then the function signature
should include an `args` parameter, i.e. be
`function(x, y, conditioning_set, suff_stat, args)`, and the additional
arguments can be passed as a list to the `args` parameter when defining the
method.

## bnlearn

bnlearn supports custom CI tests, but the interface is different from using the
engines causalDisco and pcalg. The custom test must have the signature
`function(x, y, z, data, args)`, where `x` and `y` are the variables being
tested for independence, `z` is the conditioning set, `data` is the dataset, and
`args` is a list of additional arguments. The function should return the test
statistic and the p-value. For more details on how to implement custom tests for
bnlearn, see `?bnlearn::ci.test`.

```{r}
#| label: my_custom_test_bnlearn
my_bnlearn_test <- function(x, y, z, data, args) {
  not_used <- args # Just for demonstration
  suff_stat <- list(C = cor(data), n = nrow(data))

  C <- suff_stat$C
  n <- suff_stat$n

  vars <- c(x, y, z)
  C_sub <- C[vars, vars, drop = FALSE]
  K <- solve(C_sub)

  r <- -K[1, 2] / sqrt(K[1, 1] * K[2, 2])

  z_stat <- 0.5 * log((1 + r) / (1 - r))

  stat <- sqrt(n - length(z) - 3) * z_stat

  pval <- 2 * (1 - pnorm(abs(stat)))

  c(statistic = stat, p.value = pval)
}

my_pc <- pc(
  engine = "bnlearn",
  test = my_bnlearn_test,
  alpha = 0.05,
  args = list(not_used = "Example of passing additional arguments")
)
result <- disco(data = tpc_example, method = my_pc)
plot(result)
```

# Custom Scores

Not yet implemented, but will be added in a future release. Stay tuned for
updates on this feature!
