---
title: "Knowledge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Knowledge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(causalDisco)
```

This vignette demonstrates how to use the `knowledge` function to incorporate prior knowledge into causal discovery
algorithms.
The different supported knowledge types are explained below, along with examples of how to create knowledge objects and
use them with causal discovery methods. All knowledge types can be freely combined.

At a conceptual level, all knowledge is ultimately represented as constraints on allowed edges. Some knowledge types provide higher-level abstractions for expressing common modeling assumptions more conveniently.

# Required and forbidden knowledge

At the most basic level, prior knowledge is expressed as required or forbidden edges between variables. These constraints
apply to directed edges in the causal graph.

- Required edges specify that a directed edge must exist between two variables.
- Forbidden edges specify that a directed edge is not allowed between two variables.

These constraints are specified using the %-->% (required) and %--x% (forbidden) operators.

We discussed better syntax for `%--x%`. Which of these alternatives are best:

- `child_x1 %--x% youth_x3` (current)
- `!(child_x1 %-->% youth_x3)`
- `child_x1 %!-->% youth_x3`

Note, that `child_x1 !%-->% youth_x3` is not valid syntax for the R passer.

## Specifying required and forbidden edges

Suppose we want to require an edge from A to B, and forbid an edge from B to C:

```{r required and forbidden knowledge}
kn <- knowledge(
  A %-->% B,
  B %--x% C
)
```

This knowledge object can be visualized:

```{r plot required and forbidden knowledge}
plot(kn)
```

The blue solid edge represents the required edge from A to B, while the red dashed edge represents the forbidden edge
from B to C.

## Specifying required and forbidden edges in a dataset

We will use the `tpc_example` dataset from the causalDisco package for the following examples:

```{r dataset required and forbidden knowledge}
data("tpc_example")
head(tpc_example)
```

We can pass the dataset to `knowledge`, which also checks that the specified variables exist:

```{r required and forbidden knowledge with data}
kn <- knowledge(
  tpc_example,
  child_x1 %-->% youth_x3,
  child_x2 %--x% oldage_x5
)
```

This knowledge object can also be visualized:

```{r plot required and forbidden knowledge with data}
plot(kn)
```

The plot then plots all variables in the dataset, with the required edges as blue solid edges and forbidden edges as
red dashed edges.

### Using tidyselect helpers

To make specifying variables easier, you can use tidyselect helpers such as `starts_with`:

```{r required and forbidden knowledge with tidyselect}
kn <- knowledge(
  tpc_example,
  starts_with("child") %-->% starts_with("youth"),
  starts_with("oldage") %--x% starts_with("youth")
)
```

This means, that all variables starting with "child" are required to have edges to all variables starting with "youth", and no variables starting with "oldage" can have edges to any variables starting with "youth". We can visualize this:

```{r plot required and forbidden knowledge with tidyselect}
plot(kn)
```

For a list of all available tidyselect helpers we refer to the [tidyselect reference documentation](https://tidyselect.r-lib.org/reference/index.html).

# Tier knowledge

Tier knowledge provides a higher-level abstraction for expressing systematic ordering assumptions, such as temporal or logical precedence. Internally, tier knowledge is translated into a collection of forbidden edges, but it is exposed separately because it provides a concise and structured way to express common ordering assumptions.

For example, consider a dataset with three groups of variables: child, youth, and old. We may wish to enforce that child variables precede youth variables, which in turn precede old variables. This can be expressed using tier knowledge.

Tier knowledge enforces that edges may only point from earlier tiers to later tiers. Edges within the same tier are unrestricted unless additional knowledge is supplied.

## Creating a tiered knowledge object

A tier knowledge object is created by specifying tiers and their associated variables. If numeric tiers are used, lower numbers indicate earlier tiers; otherwise, tiers are ordered by their order of appearance.

The following specifications encode the same tier structure:

```{r tier knowledge}
kn <- knowledge(
  tier(
    1 ~ c(A1, A2),
    2 ~ c(B1, B2),
    3 ~ c(C1, C2)
  )
)

# Same object, since tiers are ordered numerically
kn_same <- knowledge(
  tier(
    1 ~ c(A1, A2),
    3 ~ c(C1, C2),
    2 ~ c(B1, B2)
  )
)

# Functionally equivalent, though not identical
kn_almost <- knowledge(
  tier(
    10 ~ c(A1, A2),
    30 ~ c(C1, C2),
    20 ~ c(B1, B2)
  )
)

# Again functionally equivalent
kn_also_almost <- knowledge(
  tier(
    A ~ c(A1, A2),
    B ~ c(B1, B2),
    C ~ c(C1, C2)
  )
)
```

We can visualize the tiers using:

```{r plot tier knowledge}
plot(kn)
```

The plot then shows the tiers as layers, with the earliest tiers to the left and latest to the right.

Tidyselect helpers such as `starts_with` can also be used to define tiers in a concise way, just as with required and forbidden edges.

# Exogenous knowledge

Exogenous knowledge specifies that certain variables are exogenous, meaning they have no incoming edges in the causal graph. 
This is commonly assumed for variables that are known causes but not effects within the context of the analysis. 
Exogenous knowledge is specified using the `exogenous` function within `knowledge`.

## Specifying exogenous variables

The most natural usage is to supply the dataset so that the variables are checked for existence and selected correctly:

```{r exogenous knowledge}
kn <- knowledge(
  tpc_example,
  exogenous("child_x1")
)
```

Instead of `exogenous`, you can also use the shorthand functions `exo` or `root`.

This knowledge object can be visualized:

```{r plot exogenous knowledge}
plot(kn)
```

# Using knowledge with causal discovery

Once prior knowledge has been specified, it can be supplied to causal discovery algorithms. For example, we can use the temporal GES algorithm tges with engine "causalDisco" and temporal BIC ("tbic"):

```{r causal discovery with tier knowledge}
kn <- knowledge(
  tpc_example,
  tier(
    1 ~ starts_with("child"),
    2 ~ starts_with("youth"),
    3 ~ starts_with("oldage")
  )
)

cd_tges <- tges(engine = "causalDisco", score = "tbic")
disco_cd_tges <- disco(data = tpc_example, method = cd_tges, knowledge = kn)
```

The causal discovery algorithms respects the provided knowledge. We can plot the resulting causal graph:

```{r plot causal discovery with tier knowledge}
plot(disco_cd_tges)
```

The black edges are those inferred from the data.

# Engine specific information about knowledge

## bnlearn

All knowledge types are supported with bnlearn engines. Note, that you can get a
harmless(?) warning from bnlearn when using required knowledge.

```{r bnlearn}
data("tpc_example")

kn <- knowledge(
  tpc_example,
  child_x1 %-->% youth_x3
)

bnlearn_pc <- pc(engine = "bnlearn", test = "fisher_z", alpha = 0.05)
output <- disco(data = tpc_example, method = bnlearn_pc, knowledge = kn)
```

But the resulting causal graph respects the knowledge correctly nevertheless.

```{r plot bnlearn}
plot(output)
```

## causalDisco

WIP. Currently causalDisco only works correctly with tier and required knowledge.

## pcalg

Only forbidden knowledge symmetric knowledge is supported for pcalg. That is, edges that are forbidden in both directions.
Thus, the only type of knowledge that can be used with pcalg is knowledge created using forbidden edges (`%--x%`) without
any directed knowledge or tiers, such as this:

```{r pcalg}
data("tpc_example")
kn <- knowledge(
  tpc_example,
  child_x1 %--x% youth_x3,
  youth_x3 %--x% child_x1
)
pc_pcalg <- pc(engine = "pcalg", test = "fisher_z", alpha = 0.05)
output <- disco(data = tpc_example, method = pc_pcalg, knowledge = kn)
```

## Tetrad

WIP. Currently only implemented correctly for forbidden knowledge.



