---
title: "Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(causalDisco)
library(caugi)
```

This vignette demonstrates how to visualize causal graphs. We cover plotting `knowledge` objects and
`knowledgeable_caugi` objects, customizing layouts and styles, and exporting graphs to TikZ for inclusion in
LaTeX documents.

# Plotting Knowledge

We start by creating a simple knowledge object and visualizing it with `plot()`.
This knowledge encodes the following relationships between variables in our dataset:

```{r knowledge plot}
data(num_data)
kn <- knowledge(
  num_data,
  X1 %-->% X2,
  X2 %!-->% c(X3, Y),
  Y %!-->% Z
)

plot(kn)
```

By default, required edges are shown in blue and forbidden edges in red.
You can customize these colors using `required_col` and `forbidden_col`:

```{r knowledge plot different colors}
plot(kn, required_col = "skyblue", forbidden_col = "orange")
```

Different node layouts, node styles, and edge styles can be specified using the `layout`, `node_style`, and `edge_style`
parameters.
We provide a quick example and refer to the `causalDisco::plot()` function (and the underlying `caugi::plot()` function)
documentation for more details.

```{r knowledge plot custom styles}
plot(
  kn,
  layout = "fruchterman-reingold",
  node_style = list(
    fill = "lightblue", # Fill color
    col = "darkblue", # Border color
    lwd = 2, # Border width
    padding = 4, # Text padding (mm)
    size = 1.2 # Size multiplier
  ),
  edge_style = list(
    lwd = 1.5, # Edge width
    arrow_size = 4 # Arrow size (mm)
  )
)
```

## Plotting Tiered Knowledge

We can also visualize tiered knowledge structures. Here is an example using a dataset with variables measured at
different life stages:

```{r tiered knowledge plot}
data(tpc_example)
kn_tiered <- knowledge(
  tpc_example,
  tier(
    child ~ starts_with("child"),
    youth ~ starts_with("youth"),
    old ~ starts_with("old")
  ),
  child_x1 %-->% child_x2,
  child_x2 %!-->% youth_x3
)
plot(kn_tiered)
```

Variables are grouped by life stage, reflecting the tiered structure. This helps visually convey temporal or
hierarchical relationships in the data.

# Plotting Knowledgeable Caugi Objects

Here we create a `knowledgeable_caugi` object by combining a causal discovery result with prior knowledge,
and visualize it using the `plot` function.

```{r knowledgeable caugi plot}
data(num_data)
kn <- knowledge(
  num_data,
  X1 %-->% X2,
  X2 %!-->% c(X3, Y),
  Y %!-->% Z
)

pc_bnlearn <- pc(engine = "bnlearn", test = "fisher_z")
pc_result <- disco(num_data, method = pc_bnlearn, knowledge = kn)
plot(pc_result)
```

Blue edges indicate required relationships from prior knowledge, black edges show those learned from the data,
and forbidden edges are not plotted (to avoid confusion and clutter).

With tiered knowledge, the plotting works similarly:

```{r knowledgeable caugi tiered plot}
data(tpc_example)
kn_tiered <- knowledge(
  tpc_example,
  tier(
    child ~ starts_with("child"),
    youth ~ starts_with("youth"),
    old ~ starts_with("old")
  )
)
cd_tges <- tges(engine = "causalDisco", score = "tbic")
disco_cd_tges <- disco(data = tpc_example, method = cd_tges, knowledge = kn_tiered)
plot(disco_cd_tges)
```

Where we can see the tiers reflected in the layout of the graph.

# Exporting to TikZ

The `make_tikz()` function exports plots as clean, fully editable TikZ code for inclusion in LaTeX documents.

Unlike many graph export tools, edges are attached to nodes rather than positioned using hard-coded coordinates,
making the output easier to modify and maintain.

This functionality works for `knowledge`, `knowledgeable_caugi`, and `caugi::caugi` objects, allowing you to customize layouts,
styles, and edges further in your LaTeX document.

## Exporting Knowledge to TikZ

We first demonstrate TikZ export for a `knowledge` object.

```{r tikz export knowledge}
data(num_data)
kn <- knowledge(
  num_data,
  X1 %-->% X2,
  X2 %!-->% c(X3, Y),
  Y %!-->% Z
)

kn_plot <- plot(kn)

# Full standalone document
tikz_knowledge_code <- make_tikz(kn_plot, scale = 10, full_doc = TRUE)
cat(tikz_knowledge_code)

# Only the tikzpicture environment
tikz_knowledge_snippet <- make_tikz(kn_plot, scale = 10, full_doc = FALSE)
cat(tikz_knowledge_snippet)
```

Setting `full_doc = TRUE` generates a complete standalone LaTeX document that can be compiled directly.
Using `full_doc = FALSE` instead returns only the tikzpicture environment, which is convenient for inclusion
inside an existing LaTeX document.

The TikZ export also supports edge bending, which can substantially improve readability when edges overlap (in particular for tiered
knowledge).
This feature is not available in the standard `caugi::plot()` function. Here is an example, where the edge using the
standard straight style edge overlaps the node `youth_x3`, while the bent edge avoids this overlap:

```{r tikz bend edges}
data(tpc_example)
kn_tiered <- knowledge(
  tpc_example,
  tier(
    child ~ starts_with("child"),
    youth ~ starts_with("youth"),
    old ~ starts_with("old")
  ),
  child_x1 %-->% oldage_x5
)

kn_tiered_plot <- plot(kn_tiered)
kn_tiered_plot

tiers <- list(
  child = c("child_x1", "child_x2"),
  youth = c("youth_x3", "youth_x4"),
  old = c("oldage_x5", "oldage_x6")
)

tikz_bent_tiered <- make_tikz(kn_tiered_plot, tier_node_map = tiers, scale = 10, full_doc = FALSE, bend_edges = TRUE, bend_angle = 20)
cat(tikz_bent_tiered)
```

## Exporting Knowledgeable Caugi to TikZ

Knowledge-enhanced causal discovery results can be exported in the same way:

```{r tikz export knowledgeable caugi}
data(tpc_example)
kn_tiered <- knowledge(
  tpc_example,
  tier(
    child ~ starts_with("child"),
    youth ~ starts_with("youth"),
    old ~ starts_with("old")
  )
)

cd_tges <- tges(engine = "causalDisco", score = "tbic")
disco_cd_tges <- disco(data = tpc_example, method = cd_tges, knowledge = kn_tiered)

disco_plot <- plot(disco_cd_tges)
tikz_knowledge_snippet <- make_tikz(disco_plot, scale = 10, full_doc = FALSE)
cat(tikz_knowledge_snippet)
```

## Exporting Caugi Objects to TikZ

The same export mechanism also applies to standard caugi objects:

```{r tikz export caugi}
cg <- caugi::caugi(
  A %-->% B + C
)
plot_obj <- caugi::plot(cg, node_style = list(fill = "red"))
tikz_caugi_snippet <- make_tikz(plot_obj, scale = 10, full_doc = FALSE)
cat(tikz_caugi_snippet)
```

