---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#

The **causalDisco** causal discovery overview tool includes five simulated datasets that the code examples are run on. These five examples are:

1. `numData`: A dataset of 5 numeric variables, `X1`, `X2`, `X3`, `Z` and `Y`.
2. `catData`: A dataset of 5 categorical variables, `X1`, `X2`, `X3`, `Z` and `Y`, each taking the values `a`, `b`, `c`, `d`, `e`, `f`. 
3. `mixData`: A dataset of 5 variables, where the variables `X1`, `X2` and `X3` are identical to those from `catData`, while the variables `Z` and `Y` are identical to those from `numData`.
4. `catData_mcar`: A modification of `catData` where there has been introduced missing observations completely at random in some variables. `X1` has 10% missing values, `X2` has 5% missing values and `X3` has 20% missing values.
5. `numData_latent`: A modification of `numData` where the variable `Z` is not included and thus can be considered latent.

The code for generating the datasets is included below. Please note that the datasets do not necessarily live up to all assumptions for the proceures presented here and is only intended as example data for the syntax of the functions -- not for measuring performance. 

```{r, eval = FALSE}
############################################################################################
######################Simulate data to be used for causalDisco code examples################
############################################################################################
#All data sets use the numerical data as it offset. This dataset is simulated as 
#a mix of linear and non-linear structural equations with additive noise. The noise 
#components are Gaussian or uniform. 
#The datasets are stored in data.frames and converted to other formats (e.g. matrix) 
#in the code examples, if needed.

#sample size
n <- 1000 

#Simulate numData
set.seed(123)
numData <- data.frame(Z = abs(rnorm(n, mean = 10))) 
numData$X1 <- sqrt(numData$Z) + runif(n, min = 0, max = 2)
numData$X3 <- runif(n, min = 5, max = 10)
numData$X2 <- 2*numData$X3 - rnorm(n, mean = 5) 
numData$Y <- numData$X1^2 + numData$X2 - numData$X3 - numData$Z + rnorm(n, mean = 10)
numData <- numData[, c("X1", "X2", "X3", "Z", "Y")]

#Make catData
catData <- as.data.frame(sapply(numData, function(x) cut(x, breaks = 5,
                                                         labels = letters[1:5])))

#Make mixData
mixData <- numData
mixData$X1 <- catData$X1
mixData$X2 <- catData$X2
mixData$X3 <- catData$X3

#Make catData_mcar
catData_mcar <- catData
set.seed(1234)
catData_mcar$X1[sample(1:n, 100)] <- NA
catData_mcar$X2[sample(1:n, 50)] <- NA
catData_mcar$X3[sample(1:n, 200)] <- NA

#Make numData_latent
numData_latent <- numData[, c("X1", "X2", "Z", "Y")]

``` 