---
title: "Install `causalDisco`"
author: "Frederik Fabricius-Bjerre"
date: "2025-09-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install `pak`

We need to install `pak`, as this will help us with installing the rest of the package. 

To install `pak`, we can either use the following command:

```{r, eval=FALSE}
install.packages("pak",
  repos =
    sprintf(
      "https://r-lib.github.io/p/pak/stable/%s/%s/%s",
      .Platform$pkgType,
      R.Version()$os,
      R.Version()$arch
    )
)
```

or this

```{r, eval=FALSE}
install.packages("pak")
```

We prefer the first, as this is faster and more stable.

## Install `causalDisco`

Then, you can run 

```{r, eval=FALSE}
pak::pkg_install("github::frederikfabriciusbjerre/causalDisco")
```

This will install the latest version of `causalDisco` from GitHub. Note that there might be some system dependencies, which you might need to install first, manually. To further aid you with this, you can run

```{r, eval=FALSE}
pak::pkg_sysreqs("github::frederikfabriciusbjerre/causalDisco",
  dependencies = TRUE,
  upgrade = TRUE
)
```

This will give you a list of system dependencies, which you need to install first. Depending on your system, any system requirements for `causalDisco` might already be on your system, and this is not needed. Due to the fact that this probably requires `sudo` (admin) access to your computer, we cannot do this automatically for you.

This function _might_ not work for your system. If so, it might be favorable to look into the output of `pkg_install` and go from there. LLMs (such as ChatGPT) can be helpful in providing help with installing the packages you need for the installation of `causalDisco` to work. 

### Installing Java / JDK

If you do not have Java installed, more specifically JDK (Java Development Kit) version 25 or 21, there is a guide to installing JDK on the following link

https://github.com/cmu-phil/tetrad/wiki/Setting-up-Java-for-Tetrad

You could also use a package manager if you use Linux (or macOS with Homebrew). Here, you should search for `jdk`. 
